---
title: "First Steps in R"
output: 
  learnr::tutorial:
  progressive: true
  allow_skip: true
  df_print: default
runtime: shiny_prerendered
description: >
  Learn how to apply R to do descriptive statistics for two variables.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome

In this tutorial you get a brief introduction how to use RStudio by applying a simple R script and the console as a pocket calculator.

The following topics will be discussed

* A simple R script  
* R as a pocket calculator

## A simple R script

**Example:** Sample of heights and weights of males

$weight_i: 75,81.5,72,79.5,85.5,65,77$\
$height_i: 180,187,179,189,190,166,183$

*   Averages of weight and height  
*   Diagram of the pairs (height,weight)?

**Perform the following code**

```{r my-first-script, exercise=TRUE}
# sample: weight and height of males
weight <- c(75,81.5,72,79.5,85.5,65,77)
height <- c(180,187,179,189,190,166,183)
# means
mean(weight)
mean(height)
# diagram
plot(height,weight)

# include library
library(datasets)
data(women)
# average heights and weights for American women aged 30-39
women
```

**Exercise: Open RStudio, copy the code in the script window and execute
the code.**

> The usual Rstudio screen has four windows:

*   R scipt(s) and data view:
    +   The R script is where you keep a record of your work.
    +   script window: text editor, run button to execute marked lines
    +   data view: click on the blue circle in front of women to look at
        the dataset in a spreadsheet form.
*   Console window is the place where R is waiting for you to tell it
    what to do, and where it will show the results of a command.
    +   you can type commands and the output of executed commands
    +   R stores all commands. Using the arrows $\uparrow$ and
        $\downarrow$ you can move in the command history to repeat a
        given command.
*   Environment and history:
    +   environment tab shows all the active objects
    +   history tab shows a list of commands used so far
*   Files, plots, packages and help.
    +   files tab shows all the files and folders in your default
        workspace.
    +   The plots tab will show all your graphs.
    +   The packages tab will list a series of packages or add-ons
        needed to run certain processes: get informations about the
        package datasets
    +   For additional info see the help tab: get informations abot the
        command data and informations about women

## R as a pocket calculator

*   In the R-console you have the possibility to perform simple
    calculations and call functions.
*   Arithmetic operations are indicated by the symbols $+$, $-$, $*$,
    $/$ or \^.

**Exercise: Write the R code required to evaluate:**

* 1+1  
* $2^5$   
* 5.25/2  
* $(2+3) \cdot \sin(3)$
* $\pi$: pi  
* $\sqrt{2}$: sqrt(2) 

```{r pocket-calculator, exercise=TRUE}

```

```{r pocket-calculator-solution}
# addition
1+1
# power function
2^5
# constant
pi
# function calls
(2+3)*sin(3)
sqrt(2)
```

> Remarks:

-   Function calls are derived from the form functionname(argument1,
    argument2,... ).
    For example, you can compute $\sqrt{2}$ by the command sqrt(2), too.
-   R is case sensitive.
-   R knows a multitude of functions. In order to determine the exact
    function of a function, you can use the function help or ?. The
    syntax is help(xyz) or ?xyz.
-   If you type something in RStudio in the Script window or in the
    console, you can use the tab key to have RStudio display suitable
    commands or variable names.  
    For example, type wo or sq and press the tab key.
